---
title: "Reorientation Function"
author: "E.S."
date: "2024-04-18"
output: html_document
---




```{r}
### Reorientation of tissue positions to Fiducial Markers
library(tidyverse)
reorientation_function <- function(wp=NULL, sample=NULL, tp_file="tissue_positions.csv", mirror_row=NULL, mirror_col =NULL, mirror_row_col=NULL, plot=TRUE, save=FALSE){
  # Load data
  tp_old <- read.csv(paste0(wp, sample,"/outs/spatial/", tp_file)) 
  tp <- tp_old %>% select(1:4)
  not_to_be_changed <- tp_old %>% select(5:6)
  # Change the array
  if (mirror_col==TRUE){
  # MIRROR_COL
  # Swap in_tissue positions by array_col axis
  tp <- tp %>% add_column(., "pseudo_col" = max(.$array_col)-.$array_col, .after = 4) %>% 
  .[order(.$array_row, desc(.$array_col)),]
  # Reorder the in tissue column
  tp["in_tissue"] <- tp %>% .[order(.$array_row, desc(.$pseudo_col)), 2]
  tp$pseudo_col <- NULL
  tp <- cbind(tp, not_to_be_changed)
  } else if (mirror_row==TRUE){
    # MIRROR_ROW
  # Swap in_tissue positions by array_row axis
  tp <- tp %>% add_column(., "pseudo_row" = max(.$array_row)-.$array_row, .after = 3) %>%
  .[order(.$array_col, .$array_row, decreasing = TRUE),]
  # Reorder in tissue column
  tp["in_tissue"] <- tp %>% .[order(.$array_col, .$pseudo_row, decreasing=T), 2]
  tp$pseudo_row <- NULL
  tp <- cbind(tp, not_to_be_changed)
  }else if(mirror_row_col == TRUE){
    # MIRROR_ROW_COL
  # Swap in_tissue positions by array_col axis
  tp <- tp %>% add_column(., "pseudo_col" = max(.$array_col)-.$array_col, .after = 4) %>% 
  .[order(.$array_row, desc(.$array_col)),]
  # Reorder the in tissue column
  tp["in_tissue"] <- tp %>% .[order(.$array_row, desc(.$pseudo_col)), 2]
  tp$pseudo_col <- NULL
  # Swap in_tissue positions by array_row axis
  tp <- tp %>% add_column(., "pseudo_row" = max(.$array_row)-.$array_row, .after = 3) %>%
  .[order(.$array_col, .$array_row, decreasing = TRUE),]
  # Reorder in tissue column
  tp["in_tissue"] <- tp %>% .[order(.$array_col, .$pseudo_row, decreasing=T), 2]
  tp$pseudo_row <- NULL
  tp <- cbind(tp, not_to_be_changed)
  }
  if (plot==TRUE && all(c("ggplot2", "patchwork") %in% installed.packages())){
    library(ggplot2)
    library(patchwork)
  print(wrap_plots(tp_old %>% ggplot(aes(array_row, array_col, color=as.factor(in_tissue))) +
    geom_point() + ggtitle("Raw Spot Positions") + scale_color_manual(values=c("0"="darkgrey","1"="blue"))+ 
    theme(legend.position = "none"),
  tp %>% ggplot(aes(array_row, array_col, color=as.factor(in_tissue))) + geom_point() +
    ggtitle("Corrected Spot Positions") + scale_color_manual(values=c("0"="darkgrey","1"="blue")) +
    theme(plot.title = element_text(size = 12), legend.position = "none"), ncol = 2))
  } else {
    par(mfrow=c(1,2))
  col_tp <- rep(NA, length=nrow(tp_old[[i]]))
  col_tp[which(tp_old$in_tissue == "1")] <- "blue"
  col_tp[which(tp_old$in_tissue=="0")] <- "darkgrey"
    plot(tp_old$array_row, tp_old$array_col, col=col_tp, main="Raw Spot Positions")
  col_tp <- rep(NA, length=nrow(tp[[i]]))
  col_tp[which(tp$in_tissue == "1")] <- "blue"
  col_tp[which(tp$in_tissue=="0")] <- "darkgrey"
    plot(tp$array_row, tp$array_col, col=col_tp, main="Corrected Spot Positions")
    dev.off()
  }
  if (save==TRUE){
        ifelse(!dir.exists(paste0(wp, sample, "_corr/")), 
           dir.create(paste0(wp, sample, "_corr/")) && dir.create(paste0(wp, sample, "_corr/outs/")) && dir.create(paste0(wp, sample, "_corr/outs/spatial/")),
           print(paste0("The corrected tissue position file will be within the ", sample, "_corr/outs/spatial folder in wp")))
     write.csv(tp, paste0(wp, sample ,"_corr/outs/spatial/tissue_positions_list.csv"), row.names = FALSE) 
  }
return(tp)}
```




### Reorientation of the filtered matrices

**After** correction of tissue positions!

1) MTX Matrix
Load the raw matrix according to the Company's website [Visium: Load Matrix into R](https://www.10xgenomics.com/support/software/space-ranger/latest/analysis/outputs/space-ranger-feature-barcode-matrices), 20.04.2024

Subset the raw matrices to ens codes within the filtered matrix, using the filtered feature.tsv.gz. 
Afterwards subset the so filtered matrix to spots within the tissue using the corrected tissue positions lists.
Save the corrected filtered matrices and the new barcodes.
The feature list is similar to the feature list of the uncorrected filtered matrix, since the list of features hasn't changed.

2) H5 Matrix
To obtain the corrected h5 matrix, the corrected filtered mtx matrix will be used, to obtain data (x), indices (i) and to calculate indptr for the h5 matrix.
The uncorrected filtered h5 matrix was used, to obtain the list of features as well as the shape, since those parameters don't change, if the tissue position arrangement changes.


```{r}
library(Matrix)
matrix_correction <- function(wp=NULL, sample=NULL, new_tp_file="tissue_positions_list.csv", save=FALSE){
  # Load the corrected tissue positions
  tp <- read.csv(paste0(wp, sample, "_corr/outs/spatial/", new_tp_file)) 
  # Load the filtered genes
  filtered_feature_file <- paste0(wp, sample,"/outs/filtered_feature_bc_matrix/features.tsv.gz")
  filtered_genes <- read.delim(filtered_feature_file,header = FALSE, stringsAsFactors = FALSE)
  # Load the raw matrix
  raw_matrix_dir <- paste0(wp, sample,"/outs/raw_feature_bc_matrix/")
  matrix.raw <- paste0(raw_matrix_dir, "matrix.mtx.gz")
  barcodes.raw <- paste0(raw_matrix_dir, "barcodes.tsv.gz")
  features.raw <- paste0(raw_matrix_dir, "features.tsv.gz")
  raw_matrix <- readMM(file = matrix.raw)
  feature.names <- read.delim(features.raw, header = FALSE, stringsAsFactors = FALSE)
  barcode.names = read.delim(barcodes.raw, header = FALSE, stringsAsFactors = FALSE)
  colnames(raw_matrix) = barcode.names$V1
  rownames(raw_matrix) = feature.names$V1
  # Subset filtered_matrix
  filtered_matrix <- raw_matrix[rownames(raw_matrix) %in% filtered_genes$V1,]
  tp_in <- tp[tp$in_tissue==1,1]
  filtered_matrix <- filtered_matrix[,colnames(filtered_matrix) %in% tp_in]
  # path to the directory, where the corrected filtered mtx matrix should be saved
  corrected_filtered_matrix_dir  <- paste0(wp, sample, "_corr/outs/filtered_feature_bc_matrix/")
  # Save the corrected filtered matrix 
  matrix.new <- paste0(corrected_filtered_matrix_dir, "matrix.mtx.gz")
  barcodes.new <- paste0(corrected_filtered_matrix_dir, "barcodes.tsv.gz")
  features.new <- paste0(corrected_filtered_matrix_dir, "features.tsv.gz")
  if (save==TRUE){
    ifelse(!dir.exists(paste0(wp, sample, "_corr/")), 
           dir.create(paste0(wp, sample, "_corr/")) && dir.create(paste0(wp, sample, "_corr/outs/")),
           print(paste0("filtered_feature_bc_matrix will be within the ", sample, "_corr/outs/ folder")))
      # corrected filtered matrix as mtx.gz file
  writeMM(filtered_matrix, matrix.new)
  # corrected barcodes (colnames(filtered_matrix)) as tsv.gz file
  write_tsv(as.data.frame(colnames(filtered_matrix)),barcodes.new, col_names = FALSE)
  write_tsv(as.data.frame(rownames(filtered_matrix)),features.new, col_names = FALSE)}
return(filtered_matrix)
  }
```

h5 correction

```{r}
h5_correction <- function(wp=NULL, sample=NULL, new_tp_file="tissue_positions_list.csv",filtered_matrix_corrected=FALSE, save=FALSE){
  # Load corrected tissue positions
  tp <- read.csv(paste0(wp, sample, "_corr/outs/spatial/", new_tp_file))
  if (filtered_matrix_corrected==TRUE){
      # Load h5 matrices
  rm_h5 <- h5read(paste0(wp, sample,"/outs/raw_feature_bc_matrix.h5"), name="matrix") 
  # Load corrected filtered matrix
  matrix_dir <- paste0(wp, sample,"_corr/outs/filtered_feature_bc_matrix/")
  matrix <- paste0(matrix_dir, "matrix.mtx.gz")
  barcodes <- paste0(matrix_dir, "barcodes.tsv.gz")
  features <- paste0(matrix_dir, "features.tsv.gz")
  filtered_matrix <- readMM(file = matrix)
  feature.names <- read.delim(features, header = FALSE, stringsAsFactors = FALSE)
  barcode.names = read.delim(barcodes, header = FALSE, stringsAsFactors = FALSE)
  colnames(filtered_matrix) = barcode.names$V1
  rownames(filtered_matrix) = feature.names$V1
  # Calculate indptr
  spot_counts <- colSums(filtered_matrix>0)
  indptr <- c(0)
  for (i in 1:length(spot_counts)){
    indptr <- list.append(indptr, sum(spot_counts[1:i]))}
  # Approach without using the filtered h5 matrix
  new_filtered <- list()
  new_filtered$barcodes <- rm_h5$barcodes[rm_h5$barcodes %in% tp_in]
  new_filtered$data <- filtered_matrix@x
  new_filtered$features <- list()
  new_filtered$features$"_all_tag_keys" <- rm_h5$features$"_all_tag_keys"
  # To subset the raw h5 matrix to the features in the filtered matrix
  index <- rm_h5$features$id %in% rownames(filtered_matrix)
  new_filtered$features$feature_type <- rm_h5$features$feature_type[index]
  new_filtered$features$genome <- rm_h5$features$genome[index]
  new_filtered$features$id <- rm_h5$features$id[index]
  new_filtered$features$name <- rm_h5$features$name[index]
  new_filtered$features$target_sets <- rm_h5$features$target_sets
  new_filtered$indices <- filtered_matrix@i
  new_filtered$indptr <- indptr
  new_filtered$shape <- filtered_matrix@Dim
  }
  if (filtered_matrix_corrected==FALSE){
       # Load the filtered genes
  filtered_feature_file <- paste0(wp, sample,"/outs/filtered_feature_bc_matrix/features.tsv.gz")
  filtered_genes <- read.delim(filtered_feature_file,header = FALSE, stringsAsFactors = FALSE)
  # Load the raw matrix
  raw_matrix_dir <- paste0(wp, sample,"/outs/raw_feature_bc_matrix/")
  matrix.raw <- paste0(raw_matrix_dir, "matrix.mtx.gz")
  barcodes.raw <- paste0(raw_matrix_dir, "barcodes.tsv.gz")
  features.raw <- paste0(raw_matrix_dir, "features.tsv.gz")
  raw_matrix <- readMM(file = matrix.raw)
  feature.names <- read.delim(features.raw, header = FALSE, stringsAsFactors = FALSE)
  barcode.names = read.delim(barcodes.raw, header = FALSE, stringsAsFactors = FALSE)
  colnames(raw_matrix) = barcode.names$V1
  rownames(raw_matrix) = feature.names$V1
  # Subset filtered_matrix
  filtered_matrix <- raw_matrix[rownames(raw_matrix) %in% filtered_genes$V1,]
  tp_in <- tp[tp$in_tissue==1,1]
  filtered_matrix <- filtered_matrix[,colnames(filtered_matrix) %in% tp_in]
  # Load h5 matrices
  rm_h5 <- h5read(paste0(wp, sample,"/outs/raw_feature_bc_matrix.h5"), name="matrix") 
  fm_h5 <- h5read(paste0(wp, sample,"/outs/filtered_feature_bc_matrix.h5"), name="matrix") 
  # Calculate indptr
  spot_counts <- colSums(filtered_matrix>0)
  indptr <- c(0)
  for (i in 1:length(spot_counts)){
    indptr <- list.append(indptr, sum(spot_counts[1:i]))}
  # Approach with using the filtered h5 matrix
  new_filtered <- list()
  tp_in <- tp[tp$in_tissue==1,1]
  new_filtered$barcodes <- rm_h5$barcodes[rm_h5$barcodes %in% tp_in]
  new_filtered$data <- filtered_matrix@x
  new_filtered$features <- list()
  new_filtered$features$"_all_tag_keys" <- fm_h5$features$"_all_tag_keys"
  new_filtered$features$feature_type <- fm_h5$features$feature_type
  new_filtered$features$genome <- fm_h5$features$genome
  new_filtered$features$id <- fm_h5$features$id
  new_filtered$features$name <- fm_h5$features$name
  new_filtered$features$target_sets <- fm_h5$features$target_sets
  new_filtered$indices <- filtered_matrix@i
  new_filtered$indptr <- indptr
  new_filtered$shape <- fm_h5$shape
  }
  if (save==TRUE){
  H5Fcreate(paste0(wp, sample, "_corr/outs/filtered_feature_bc_matrix.h5"))
  h5closeAll()
  h5write(new_filtered, paste0(wp, sample, "_corr/outs/filtered_feature_bc_matrix.h5"), name="matrix")
  }
  return(new_filtered)
  }
```



### Control

To check whether the spot reorientation produced a correct orientation of the spots, the number of total counts for each spot can be compared, using the uncorrected vs. the corrected array positions. 
To align the array positions with the tissue position, $max(array|row)-array|row$ has been used.

```{r}
control_reorientation <- function(wp=NULL, sample=NULL,  tp_file="tissue_positions_list.csv", new_tp_file="tissue_positions_list.csv"){
  # Load uncorrected and corrected tissue positions and the low resolution image
  tp_old <- read.csv(paste0(wp, sample, "/outs/spatial/", tp_file)) 
  tp_new <- read.csv(paste0(wp, sample, "_corr/outs/spatial/", new_tp_file))
  # Combine the uncorrected vs. corrected localization within the array
  tp_compared <- merge(tp_old[, 1:4], tp_new[,1:4], 
                by=c("barcode", "array_row", "array_col"), suffixes=c("_uncorr", "_corr"))
  tp_compared <- cbind(tp_compared, "in_tissue_compared" = rowSums(tp_compared[,
                c("in_tissue_uncorr", "in_tissue_corr")]))
  tissue_list <- c()
  for (row in 1:nrow(tp_compared))
    tissue_list <- list.append(tissue_list, if(tp_compared$in_tissue_compared[row] == 1 &
                                               tp_compared$in_tissue_uncorr[row] == 1){
      "U"
    }else if((tp_compared$in_tissue_compared[row] == 1 & tp_compared$in_tissue_corr[row] == 1)){
      "C"
    }else{
     tp_compared$in_tissue_compared[row]})
  tp_compared["in_tissue_compared"] <- tissue_list
  # Load the raw matrix
  raw_matrix_dir <- paste0(wp, sample,"/outs/raw_feature_bc_matrix/")
  matrix.raw <- paste0(raw_matrix_dir, "matrix.mtx.gz")
  barcodes.raw <- paste0(raw_matrix_dir, "barcodes.tsv.gz")
  features.raw <- paste0(raw_matrix_dir, "features.tsv.gz")
  # load the raw matrix.mtx.gz
  raw_matrix <- readMM(file = matrix.raw)
  # load the raw feature.tsv.gz
  feature.names <- read.delim(features.raw, header = FALSE, stringsAsFactors = FALSE)
  # load the raw barcodes.tsv.gz
  barcode.names = read.delim(barcodes.raw, header = FALSE, stringsAsFactors = FALSE)
  # set colnumn names to the barcodes in the array                         
  colnames(raw_matrix) = barcode.names$V1
  # and the row names to the raw features of the array   
  rownames(raw_matrix) = feature.names$V1
  # Calculate Counts per Spot (CPS)
  counts_per_spot <- data.frame("barcode" = colnames(raw_matrix), "cps" = colSums(raw_matrix))
  cps_old <- merge(tp_old, counts_per_spot, by="barcode")
  cps_new <- merge(tp_new, counts_per_spot, by="barcode")
  rows <- 0:77
  columns <- 0:127
  # Comparison of TP and CPS
  # TP
  plot_tp <- tp_compared %>% ggplot(aes(max(array_row)-array_row, array_col, color= as.factor(in_tissue_compared))) + geom_point(size=2) + ggtitle(paste0("TP Compared")) + scale_color_manual(values=c("0"="darkgrey","2"="gold1","U"="darkorange","C"="blue"), name="Spot", labels=c("0"="Not in Both","2"="In Both","U"="Only uncorrected","C"="Only corrected")) + theme(legend.text=element_text(size=8))
  # CPS uncorrected
  plot_uncorr <- cps_old %>% filter(in_tissue==1) %>% ggplot(aes(max(array_row)-array_row, array_col, color = cps)) + geom_point() + ggtitle("CPS: TP_uncorrected") + scale_colour_gradientn(colours = c("blue", "white", "yellow"))
  # CPS corrected
plot_corr <- cps_new %>% filter(in_tissue==1) %>% ggplot(aes(max(array_row)-array_row, array_col, color = cps)) + geom_point() + ggtitle("CPS: TP_corrected") + scale_colour_gradientn(colours = c("blue", "white", "yellow"))
compare_tp <- grid.arrange(plot_image, plot_tp,  plot_uncorr, plot_corr, ncol=2)
print(compare_tp)
# Save graph
#  ggsave(paste0(wp, sample, "_corr/outs/spatial/compare_tp_uncorr_corr.png"), plot = compare_tp, width =18, height = 12, units = "cm")
}

```

As seen in the plots, spots with a high number of counts correlate with spots assigned to be in the tissue (`1`) as well as with histological structures within the tissue, which undergirds also that the barcodes within the array were captured at correct barcode positions to align with the image, if the symbols were correct detected.





